---
title: "README"
author: "JMV"
date: "1/21/2022"
output: github_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Project Description

This document describes the project for the final project for the "Getting and Cleaning Data" course on Coursera.

From the course site:

"The purpose of this project is to demonstrate your ability to collect, work with, and clean a data set. The goal is to prepare tidy data that can be used for later analysis. You will be graded by your peers on a series of yes/no questions related to the project. You will be required to submit: 1) a tidy data set as described below, 2) a link to a Github repository with your script for performing the analysis, and 3) a code book that describes the variables, the data, and any transformations or work that you performed to clean up the data called CodeBook.md. You should also include a README.md in the repo with your scripts. This repo explains how all of the scripts work and how they are connected."

"One of the most exciting areas in all of data science right now is wearable computing. Companies like Fitbit, Nike, and Jawbone Up are racing to develop the most advanced algorithms to attract new users. The data linked to from the course website represent data collected from the accelerometers from the Samsung Galaxy S smartphone. A full description is available at the site where the data was obtained:

[Link to description of how dataset was obtained.](http://archive.ics.uci.edu/ml/datasets/Human+Activity+Recognition+Using+Smartphones)

[Link to download dataset for project](https://d396qusza40orc.cloudfront.net/getdata%2Fprojectfiles%2FUCI%20HAR%20Dataset.zip)

You should create one R script called run_analysis.R that does the following. 

1) Merges the training and the test sets to create one data set.

2) Extracts only the measurements on the mean and standard deviation for each measurement. 

3) Uses descriptive activity names to name the activities in the data set

4) Appropriately labels the data set with descriptive variable names. 

5) From the data set in step 4, creates a second, independent tidy data set with the average of each variable for each activity and each subject.


## Description of analyis files

Descriptions of the experiment set up and files are found in the README.txt file. The R script `run_analysis.R` was used to combine the measurements from test and training subjects into one dataset. The original datasets contained 561 features that were extracted from the accelerometer measurements (see description of these in the README.txt and features_info.txt files). During processing with `run_analysis.R`, a subset of these features (the mean and standard deviation) were retained. The `run_analysis.R` script outputs a tidy datset `tidy_samsung_accelerometer_data.csv`, that summarises each of the mean/standard deviation (std) values for each of the 30 subjects in the study, for each of the six activities for which measurements were obtained. 

As shown in the table below, column names correspond to measured variables, with variables begining with 't' representing measurements in the time domain, and those beginning with 'f' representing Fourier transformed values. 

```{r echo=FALSE}
df <- read.csv("./UCI HAR Dataset/tidy_samsung_accelerometer_data.csv")
knitr::kable(df[1:10, 2:10])
```



